Trade-offs involving juvenile competition offer more robust coexistence than a fecundity-defence trade-off in communities dependent on disturbance for coexistence

### 2014-02-04 MSv1.tex
new model sectio etcn

\section{The Model}
\label{model}
We consider a simple lottery type model, originally introduced by \cite{sale1978coexistence} and \cite{chesson1981environmental} and used extensively since \citep[e.g.][]{muko2000species,pacala1992herbivores}. There are a fixed number $N$ distinct sites, each capable of supporting a single adult individual, observed in continuous time $\tau$. At discrete times $\tau_1,\tau_2, \dots$, there is a tree death, selected at random from the community, followed by competition between the remaining individuals for the vacated site. We assume that the time to fill a gap with an individual is small enough that the gap created at time $\tau_t$ is filled by an adult individual by time $\tau_{t+1}$, i.e. death-replacement events are sufficiently far apart temporally that they do not overlap and can be considered as discrete events. This assumption makes it possible to model this system as a time homogeneous Markov Chain (MC) model. With the range of parameters considered below, the expected gap replenishment time is approximately 4 years. Simulations reducing the number of adult trees by 4\% (consistent with 1\% annual mortality for this replenishment period) do not exhibit any qualitative difference.

Gaps are occupied by populations of two species $N_1(t), N_2(t)$, such that $N_1(t) + N_2(t) = N$, where $t$ represents time as measured by the number of replacement events that have occurred, i.e. $N_i(t)$ represents the population of species $i$ after the gap created at time $\tau_t$ has been filled, and before another gap is created at time $\tau_{t+1}$. Because after each death-replacement event, the number of individuals is fixed, we consider the dynamics of species 1 only, modelling $N_1(t)=n$ with $N_2(t)=N-n$. We assume 

@@

\item The system is closed, so no seeds are introduced from elsewhere. Biologically, this could suggest an island population or community isolated by other geographic constraints. A low level of immigrating seeds does not qualitatively affect the results.
\item The individuals considered are asexual or self-fertilising, so that a single individual can produce seeds, and that individuals are always reproductively active.
\item The two species differ only in growth rates $g_i$, annual per capita seed production $s_i$ and resistance to disturbance $I_i$, with pairwise trade-offs between the three traits considered. Species 2 is assumed to have the higher growth rate of the two species, $g_2\geq g_1$. Note that if no trade-off occurs, coexistence is impossible. When species are identical, the system follows a random walk until one species goes extinct, while if one species dominates one of more life history trait, this species will exclude the other.
\item As individuals compete for light in a gap, we assume complete asymmetric competition. Once the combined size of the juveniles in a gap is that of a single adult, then the largest individual will win the site. Competition for a site is therefore a race to reach size $C/2$, where $C=100$ is the size of an adult individual.
\item It is possible for the faster growing species 2 to usurp juveniles of the slower growing species. The initial juvenile to reach a gap is determined by lottery, but species 2 may take over the site if further seeds arrive before a species 1 juvenile has grown sufficiently large. To usurp species 1, species 2 requires seeds to arrive before species 1 has been present for time $x=C(1/g_2-1/g_1)/2$.
\item Seeds are all assumed to be viable and are dispersed at random in the environment by a homogeneous Poisson process (with rate $\lambda_i =s_iN_i(t)/N$) as per high distance dispersals in \cite{clark1999interpreting}.
\item Disturbance events occur at each time step with probability $f=(dNT_D)^{-1}$ where $T_D$ is the expected number of years between disturbance events, and $d$ is the expected annual mortality rate (fixed at $1\%$). Within these events, individuals die with species-specific probability $I_i$ (a composite measure of disturbance intensity and species resistance), before the remaining individuals compete to fill the emptied sites. Increasing either frequency $f$ or intensity $I_i$ will result in an increase in total basal area lost over a given time frame, a standard measure of disturbance \citep[e.g.][]{molino2001tree,peterson1997tornado}.

@@

When a disturbance does not occur (with probability $1-f$), the number of species 1 individuals will increase if a species 2 individual dies, seeds from species 1 reach the site first, and then grow for time $x$ without species 2 reaching the site. As both species have the same dispersal kernel, this gives

@@

P(\text{Increase}(n))&=P(N_1(1)=n+1)|N_1(0)=n) \\

@@

where the exponential term is the probability that no species 2 seeds arrive in time $x$. Similarly, species 1 will decline in numbers if it suffers a death, and species 2 claims the site, either arriving at the site first or by successfully invading a site where species 1 juveniles are present, giving

@@

P(\text{Decrease}(n))=&P(N_1(1)=n-1 | N_1(0)=n)  \\ 

@@

All other possibilities result in the populations of the two species remaining the same after a time step, so we use a third function $P(\text{Stay}(n))=1-P(\text{Increase}(n))-P(\text{Decrease}(n))$ to fully define the transition probabilities, giving a $(N+1) \times (N+1)$ tridiagonal transition matrix $A$, with $A_{i,i}=P(\text{Stay}(i-1)), A_{i,i+1}=P(\text{Increase}(i-1)), A_{i,i-1}=P(\text{Decrease}(i-1))$.

During a disturbance event, each individual of species $i$ has a probability $I_i$ of mortality, resulting in $d_{dist}=d_1+d_2$ total deaths, where $d_i$ is the number of species $i$ deaths. The remaining $N-d_{dist}$ individuals compete over the empty site, with species 1 colonising if it reaches the site at least time $x$ before species 2. Therefore, species 1 will take a site with a probability defined by the function
\begin{equation}
\label{sp1}
Sp_1(n_1^*,n_2^*)=\frac{s_1 n_1^*}{s_1n_1^*+s_2n_2^*}\exp \left(-s_2 x\frac{n_2^*}{N}\right),
\end{equation}
where $n_1^*,n_2^*$ are the populations remaining immediately after the deaths from a disturbance event.

To determine system coexistence, we use stochastic invasion analysis, similar to that introduced by \cite{chesson1989invasibility} and \cite{ellner1989convergence}. \cite{chesson1982stabilizing} proved that when the boundary growth rates, or average changes at the boundaries are positive, such that each species will on average increase when rare, the populations are stochastically bounded away from zero, and coexistence is expected.  Note that because the model presented here uses a discrete state space, the invasion criteria are considered with a single individual of the invader present, as opposed to being assessed at 0 as in a continuous state space (we are modelling individuals, not densities or proportions of sites occupied). We approximate the expected change at the boundaries ($n=1,N-1$) using the formula
\begin{equation}
\label{avchange}
\begin{array}{rl}
\text{AverageChange}(n) = (1-f)\left[P(\text{Increase}(n))-P(\text{Decrease}(n))\right]&\\[0.75em]
+ f\left[ (nI_1 + (N-n)I_2) Sp_1(n(1-I_1),(N-n)(1-I_2)) - nI\right]&.
\end{array}
\end{equation}
The first term in \eqref{avchange} is the expected change when disturbance does not occur multiplied by the within-event probability that there is no disturbance event. The second term indicates the expected change during a disturbance as measured by the expected remaining populations. This only approximates the actual expected average change, yet is a good approximation (Appendix~\ref{app:approximations}). Coexistence is predicted for regions of parameter space where the following 2 inequalities both hold;
\begin{align}
\label{ac1}\text{AverageChange}(1)&>0, \\
\label{acn-1}\text{AverageChange}(N-1)&<0. \end{align}
Note that when there is no disturbance ($T_D=\infty$), these conditions simplify to 
\begin{align}
\label{lowerboundarycond}P(\text{Increase}(1))&>P(\text{Decrease}(1)) \\
\label{upperboundarycond}P(\text{Increase}(N-1))&<P(\text{Decrease}(N-1)). \end{align}
In biological terms, these conditions can be considered as the following: If there is a single individual of a species remaining, for that species to persist it must produce offspring that successfully colonise at least one other site before the adult dies. We implicitly assume that either individuals are asexual, or that self-fertilisation is possible, and more complex models are required to investigate sexual populations that are self-incompatible. The colonisation of a second site therefore must be more likely than the death of this single individual. Analyses using these conditions are extensively checked against time series simulations.

@@

\section*{Acknowledgements}
This work was funded by a NERC studentship for SN.

@@

 &+ \left(1-\frac{1}{dNT_D}\right)(P(\text{Increase}(1))-P(\text{Decrease}(1))), \notag

@@

 &+ \left(1-\frac{1}{dNT_D}\right)(P(\text{Increase}(N-1))-P(\text{Decrease}(N-1))). \notag

@@

>>>>

\section{The Model}
\label{model}
We consider a lattice of a number $M$ of fixed sites with size $C$. Each site can contain one adult individual of one of two species present in the community, or in the absence of an adult individual, may contain juveniles. These sites are observed in continuous time, and at times $\tau_1,\tau_2,\dots$ there is an adult death, with the individual randomly selected from a uniform distribution, such that each adult present in the community has an equal chance of mortality. These death events are followed by competition among juveniles to claim the site that previously contained the dead adult. We assume

@@

\item The death of adults and the growth of successful juveniles occur in a ratio of 1:1, such that the number of adults present $N$ is constant when considered at times $\tau_t,t=1,2,\dots$. This enables us to consider only the population dynamics of one species, with the population of the second species determined by the relationship $N_1(t) + N_2(t) = N$, where $t$ represents time as measured by the number of death events that have occurred, and $N_i(t)$ represents the population of species $i$ is the population after the $t$-th death event. We therefore consider the dynamics only of the species 1 population $n$, with $N_1(t)=n, N_2(t)=N-n$.
\item The two species considered are identical in all bar two aspects. They differ only in annual per capita seed production $s_i$ and juvenile growth rates $g_i$. These juvenile growth rates are considered to be a linear growth from a seed of size $0$ to an adult occupying the entire site, i.e. of size $C$. We assume without loss of generality species 2 is the faster growing species, $g_2>g_1$.
\item Both species are assumed to be asexual or self fertilizing, such that a single individual can produce seeds and therefore offspring.
\item Following the death of an adult individual, the remaining $N-1$ individuals produce seeds that can colonise the empty site, or gap. Seeds cannot survive in the understorey, and die immediately. Therefore, when an adult dies, the gap created is considered to be empty until new seeds arrive at that site.
\item All seeds that arrive in a gap are considered viable, and in the absence of competition would successfully reach adulthood. They are produced in time according to a Poisson process with $\lambda_i =  s_i N_i$ the average number of seeds to be produced per unit time.
\item The seeds produced are spread uniformally between sites. After a death event, the expected number of seeds from species $i$ to arrive per unit time at the gap created is given by $s_i \hat{N}_i /N$, where $\sum_{i=1}^2 \hat{N}_i =N-1$ due to the death of an individual.
\item Once seeds arrive in an site not containing any adults, they instantly begin to grow at rate $g_i$, there is no latency period. When all the juveniles in a given site combine to occupy the entire gap, that is when the sum of their sizes is equal to $C$, then the smallest of these juveniles will die while the others continue to grow. This continues until there is only a single individual remaining, which will then occupy the site to adulthood. Note that since there is no intraspecific variation, this allows us to only consider the first seed of each species to arrive at a site.
\item Only considering the first seeds of each species to arrive in a gap, colonisation of a gap becomes a race to colonise and grow to size $C/2$, and any other individuals present will have size less than $C/2$ and will therefore be killed when in competition wit the larger juvenile.

@@

These assumptions allow us to model the system as a time homogeneous Markov Chain model where states are defined by the number of species 1 adults $n$, each time step is the replacement of an individual adult that dies, and the replacement of that individual is determined by the present populations of each species present in the community. This, the number of species 1 individuals $n$ will increase if an individual of species 2 dies, species 1 seeds arrive first and reach size $C/2$ before any subsequent species 2 seeds that arrive in the site reach that height. A seed of species 2 cannot reach the height $C/2$ before species 1 if it arrives in the site when species 1 has been present for a time greater than $x=C(1/g_2-1/g_1)/2$. As both species distribute seeds randomly throughout the environment, the probability of species 1 increasing in number is given by

@@

p_{n,n+1}=&P(N_1(1)=n+1)|N_1(0)=n) \\

@@

where the exponential term is the probability that no seeds of species 2 arrive in time $x$. Similarly, species 1 will decline in numbers if it suffers a death, and species 2 claims the site, either arriving at the site first or by successfully invading a site where species 1 juveniles are present, giving

@@

p_{N-1,N}=&P(N_1(1)=n-1 | N_1(0)=n)  \\ 

@@

All other possibilities result in the populations of the two species remaining the same after a time step, so we use a third function $p{n,n}=1-p_{n,n+1}-p_{n,n-1}$ to fully define the transition probabilities, giving a $(N+1) \times (N+1)$ tridiagonal transition matrix $A$, with $A_{i,i}=p_{n,n}, A_{i,i+1}=p_{n,n+1}, A_{i,i-1}=p_{n,n-1}$. Note that these rules ensure the system is closed, with no seeds arrived from outside the system, there therefore that the probability of moving from a state where a species is absent to a state where it is present equals zero. Mathematically, this gives a system with absorbing boundaries when either species is extinct; $n \in \{0,N\}$.

We also consider disturbance events that occur during the time step $t \to t+1$ with probability $f=(dNT_D)^{-1}$ where $T_D$ is the expected number of years between disturbance events, and $d$ is the expected annual mortality rate (fixed at $1\%$). Within these events, individuals die with probability $I$ (the intensity of the disturbance), before the remaining individuals compete to fill the emptied sites. Increasing either frequency $f$ or intensity $I$ will result in an increase in total basal area lost over a given time frame, a standard measure of disturbance \citep[e.g.][]{molino2001tree,peterson1997tornado}. During a disturbance event, each individual has a probability $I$ of mortality, resulting in $d_{dist}=d_1+d_2$ total deaths, where $d_i$ is the number of species $i$ deaths. The remaining $N-d_{dist}$ individuals compete over the empty site, with species 1 colonising if it reaches the site at least time $x$ before species 2. Therefore, species 1 will take a site with a probability defined by the function
\begin{equation}
\label{sp1}
Sp_1(n_1^*,n_2^*)=\frac{s_1 n_1^*}{s_1n_1^*+s_2n_2^*}\exp \left(-s_2 x\frac{n_2^*}{N}\right),
\end{equation}
where $n_1^*,n_2^*$ are the populations remaining immediately after the deaths from a disturbance event.

We mostly consider the case where each adult individual is reproductively active at all times, but in real communities, the proportion of adults that are reproductively active can vary significantly. Therefore, we also consider the case where in each time step, each individual is producing seeds with probability $p$, which has the effect of reducing the productivity of the system. This parameter $p$ is assumed to be time invariant and independent of species abundance, as well as constant across species. This allows us to consider the effects of seed limitation, which we compare with other ways of altering productivity.

\section{Conditions for Coexistence}
To determine whether two species coexistent in the model, we use a version of stochastic boundedness. Chesson (1982) demonstrated that if the average growth rates at the boundaries of a lottery model with infinite sites are positive, the model is stochastically bounded, and both species coexist. We therefore consider the expected change when a species is extremely rare, with a single individual present (i.e. $n=1,N-1$), which we approximate using the function
\begin{equation}
\label{avchange}
\begin{array}{rl}
\text{AverageChange}(n) = (1-f)\left[p_{n,n+1}-p_{n,n-1}\right]&\\[0.75em]
+ f\left[ NI Sp_1(n(1-I),(N-n)(1-I)) - nI\right]&.
\end{array}
\end{equation}
The first term in \eqref{avchange} is the expected change when disturbance does not occur multiplied by the within-event probability that there is no disturbance event. The second term indicates the expected change during a disturbance as measured by the expected remaining populations. Due to Jensen's inequality, this is not the actual expected change, but an approximation. This approximation improves as system capacity $N$ is increased, as demonstrated in Appendix~\ref{app2a}, and works well for $N\geq500$. Coexistence is predicted for regions of parameter space where the following 2 inequalities both hold;
\begin{align}
\label{ac1}\text{AverageChange}(1)&>0, \\
\label{acn-1}\text{AverageChange}(N-1)&<0. \end{align}
Note that when there is no disturbance ($T_D=\infty$), these conditions simplify to 
\begin{align}
\label{lowerboundarycond}p_{1,2} &>p_{1,0} \\
\label{upperboundarycond}p_{N-1,N} &<p_{N-1,N-2}. \end{align}
In biological terms, these conditions can be considered as the following: If there is a single individual of a species remaining, for that species to persist it must produce offspring that successfully colonise at least one other site before the adult dies. We show in Figure~\ref{confidenceints} and Appendix~\ref{app:stationary} that this is a good approximation of coexistence observed within time series simulations.

Unlike the work of Chesson (1982), we are modelling finite populations, rather than considering the proportion of sites occupied by a species. Therefore,  the boundary conditions are considered with a single individual present, rather than being evaluated at 0 as in Chesson's work. This combined with the non zero probability of a single individual dying at a given time ensures that the only stable distribution for the model as $t \to \infty$ is $P(N_1(t)=0) = P(N_1(t)=N) = 1/2.$ That is, the coexistence predicted by stochastic boundedness in a closed system is unstable. Despite this, as demonstrated by Figure~\ref{confidenceints}, times series simulations do demonstrate coexistence on ecological time scales. Further, as system capacity $N$ increases, the model is well approximated by considering the proportion of sites $Q_i$ containing species $i$. Then, performing an analysis as in Chesson (1982) we can return the conditions for coexistence given above when considering the asymptotic growth rate of a species with zero density. At lower system capacity, a low level of immigration from elsewhere can give a fixed distribution where species can coexist, and as we show in Appendix~\ref{app:stationary}, a very low level of immigration produces a fixed, stable distribution that is maximised at an internal point $x \in [2,N]$ such that $\pi_x > \pi_{1}, \pi_{N+1}$ where $\pi_i$ is the probability of having $i-1$ species 1 individuals at the stable distribution) with approximately the same conditions as the stochastic boundedness criteria outlined above.

\begin{figure}[htbp]
\begin{center}
\begin{tabular}{cccc}
(a)&&(b)&\\
&\includegraphics[width=2.5in]{lowxCIplot.pdf} && \includegraphics[width=2.5in]{medxCIplot.pdf}\\
(c)&&&\\
&\includegraphics[width=2.5in]{highxCIplot.pdf}&& \end{tabular}
\caption{(a) Time series simulations satisfying the conditions \eqref{lowerboundarycond} but not \eqref{upperboundarycond} (x=0.01) show extinction of the more fecund species 1. (b) When both conditions are satisfied (x=0.035), both species will coexist in the environment over ecological time scales. (c) When \eqref{upperboundarycond} is satisfied yet \eqref{lowerboundarycond} is false (x=0.06), species 2 will go extinct. Lines show the mean and 95\% confidence intervals of 1000 simulations in each scenario. Parameters $s_1=500,s+2=50, N=1000$.}
\label{confidenceints}
\end{center}
\end{figure}

@@

\section*{Acknowledgements}
This work was funded by a NERC studentship for SN. Thanks go to Rob Beardmore and Ben Collen for comments on an earlier version of this manuscript.

@@

 &+ \left(1-\frac{1}{dNT_D}\right)(p_{1,2}-p_{1,0}), \notag

@@

 &+ \left(1-\frac{1}{dNT_D}\right)(p_{N-1,N}-p_{N-1,N-2}). \notag

@@

   \section{Stationary distributions, and the effects of low level immigration}
\label{app:stationary}
In a closed system such as the one detailed above, if there is a non zero probability of reaching a state from which there is zero probability of leaving, the only stable distribution is fixed at one of those absorbing states. On the model considered in the current manuscript, the result of this is that the stable distribution $\Pi$ is given by
\begin{equation}
\pi_i = \begin{cases}
1/2 , i = 0,N \\
0, i = 1,2,\dots,N-1. \end{cases}
\end{equation}
Therefore, any coexistence suggested by invasion analysis is merely temporary. Over ecological timescales of decades or centuries, both species may coexist over several generations,yet at time $t$ tends to infinity, one species will go extinct, although the identity of that species is unknown.

However, when low levels of immigration from outside the system is included, studying the stationary distribution of a system without disturbance demonstrates that the stochastic boundedness conditions are a good approximation of where coexistence is expected to occur (here coexistence is considered to occur when the most probably state in the stationary distribution has both species present).

Compared to this definition of coexistence, we find that stochastic boundedness does slightly enlarge the expected region of coexistence, but generally captures the expected behaviour well. With parameters $s_1=500,s_2=50,N=1000$, we find that when only 10 seeds of each species are expected to arrive in the system each time step, stochastic boundedness conditions adjusted to include this low level of immigration give an expected range of coexistence for $0.015 \lessapprox x \gtrapprox 0.046$.

The stationary distribution is found by solving
\begin{equation}
A \mathbf{\pi} = \mathbf{\pi}
\end{equation}
where $\mathbf{\pi}$ is the stationary distribution such that $\pi_i$ is the probability of having $i-1$ species 1 individuals present.

The range of $x$ which have an interior maximum is given by $0.02 \lessapprox x \gtrapprox 0.041$. If we consider parameters where an interior peak is present ($\pi_x > \pi_{x-1},\pi_{x+1}$ for some $x \in [1, N-1]$), this range increases to  although the most likely state is still extinction of one species $0.019 \lessapprox x \gtrapprox 0.043$ although at the extremes of this range, the most likely state the system occupies is extinction of one species. This slight overestimation of range by the stochastic boundedness conditions is consistent across parameter changes, Combining this observation with the time series simulations in Figure~\ref{confidenceints}  we conclude that it is an appropriate approximation of the systems behaviour, over ecological timescales in the absence of immigration, and the behaviour as $t \to \infty$ when immigration is considered.
